---
title: "Assignment #1"
format: html
---

## Assignment 1

You only need to write lines of code for each question. When answering questions that ask you to identify or interpret something, the length of your response doesn’t matter. For example, if the answer is just ‘yes,’ ‘no,’ or a number, you can just give that answer without adding anything else.

We will go through comparable code and concepts in the live learning session. If you run into trouble, start by using the help help() function in R, to get information about the datasets and function in question. The internet is also a great resource when coding (though note that no outside searches are required by the assignment!). If you do incorporate code from the internet, please cite the source within your code (providing a URL is sufficient).

Please bring questions that you cannot work out on your own to office hours, work periods or share with your peers on Slack. We will work with you through the issue.

You will need to install PLINK and run the analyses. Please follow the OS-specific setup guide in [`SETUP.md`](../../SETUP.md). PLINK is a free, open-source whole genome association analysis toolset, designed to perform a range of basic, large-scale analyses in a computationally efficient manner.

```{r setup, include=FALSE}
# Set up a consistent path for all chunks of code
knitr::opts_knit$set(root.dir = normalizePath("../../"))
```

```{r, message=FALSE, warning=FALSE, results='hide'}
library(data.table)
library(ggplot2)
library(seqminer)
library(HardyWeinberg)
library(dplyr)
```

#### Question 1: Data inspection

Before fitting any models, it is essential to understand the data. Use R or bash code to answer the following questions about the `gwa.qc.A1.fam`, `gwa.qc.A1.bim`, and `gwa.qc.A1.bed` files, available at the following Google Drive link: <https://drive.google.com/drive/folders/11meVqGCY5yAyI1fh-fAlMEXQt0VmRGuz?usp=drive_link>. Please download all three files from this link and place them in `02_activities/data/`.


(i) Read the .fam file. How many samples does the dataset contain?

```{bash}         
wc -l ./02_activities/data/gwa.qc.A1.fam
```

The dataset contains 4000 samples

(ii) What is the 'variable type' of the response variable (i.e.Continuous or binary)?

Continuous

(iii) Read the .bim file. How many SNPs does the dataset contain?

```{bash}         
wc -l ./02_activities/data/gwa.qc.A1.bim
```

The dataset contains 101083 SNPs

#### Question 2: Allele Frequency Estimation

(i) Load the genotype matrix for SNPs rs1861, rs3813199, rs3128342, and rs11804831 using additive coding. What are the allele frequencies (AFs) for these four SNPs?

```{bash,message=FALSE, warning=FALSE}
setopt interactivecomments
printf "%s\n" rs1861 rs3813199 rs3128342 rs11804831 > ./02_activities/data/snplist.A1.txt

plink2 --bfile ./02_activities/data/gwa.qc.A1 \
      --extract ./02_activities/data/snplist.A1.txt \
      --freq \
      --out ./02_activities/data/geno.additive.A1.freq
```

```{r,message=FALSE, warning=FALSE}
freq.A1 <- fread("./02_activities/data/geno.additive.A1.freq.afreq")
head(freq.A1)
```

The allele frequencies for the 4 SNPs are:
rs1861: 0.0539859
rs3813199: 0.0569126
rs3128342: 0.3051210 
rs11804831: 0.1543410

(ii) What are the minor allele frequencies (MAFs) for these four SNPs?

The MAF for the 4 SNPs are:
rs1861: 0.0539859
rs3813199: 0.0569126
rs3128342: 0.3051210 
rs11804831: 0.1543410
All are below 0.5 and therefore minor alleles

#### Question 3: Hardy–Weinberg Equilibrium (HWE) Test

(i) Conduct the Hardy–Weinberg Equilibrium (HWE) test for all SNPs in the .bim file. Then, load the file containing the HWE p-value results and display the first few rows of the resulting data frame.

```{bash}    
plink2 --bfile ./02_activities/data/gwa.qc.A1 --hardy --out ./02_activities/data/gwa_qc_hwe_A1
```

```{r,message=FALSE, warning=FALSE}    
hwe.A1 <- fread("./02_activities/data/gwa_qc_hwe_A1.hardy")
head(hwe.A1)
```

(ii) What are the HWE p-values for SNPs rs1861, rs3813199, rs3128342, and rs11804831?

```{r,message=FALSE, warning=FALSE}  
hwe.A1[hwe.A1$ID == "rs1861", "P"]
hwe.A1[hwe.A1$ID == "rs3813199", "P"]
hwe.A1[hwe.A1$ID == "rs3128342", "P"]
hwe.A1[hwe.A1$ID == "rs11804831", "P"]
```

The p-values for the 4 SNPs are:
rs1861: 0.274719
rs3813199: 1
rs3128342: 0.330273 
rs11804831: 0.113354

#### Question 4: Genetic Association Test

(i) Conduct a linear regression to test the association between SNP rs1861 and the phenotype. What is the p-value?

```{r,message=FALSE, warning=FALSE}
geno_subset_A1 <- fread("./02_activities/data/geno.additive.A1.raw")
geno_subset_A1 = na.omit(geno_subset_A1)

model.A1 <- glm(PHENOTYPE ~ rs1861_C, data = geno_subset_A1, family = gaussian)
summary(model.A1)
```

The p-value is < 2e-16

(ii) How would you interpret the beta coefficient from this regression?

The regression coefficient for rs1861 is 0.96698 (p < 2e-16), indicating that each additional copy of the C allele is associated with an increase of approximately 0.967 units in the phenotype on average

(iii) Plot the scatterplot of phenotype versus the genotype of SNP rs1861. Add the regression line to the plot.

```{r, message=FALSE, warning=FALSE}         

 geno_range_A1 <- data.frame(
   rs1861_C = seq(0, 2, length.out = 100)
 )

 geno_range_A1$predicted_prob_A1 <- predict(model.A1, newdata = geno_range_A1, type = "response")

count_data <- geno_subset_A1 %>%
  filter(!is.na(rs1861_C), !is.na(PHENOTYPE)) %>%
  group_by(rs1861_C, PHENOTYPE) %>%
  summarise(count = n(), .groups = "drop")

p1=ggplot() +
  geom_point(
    data = count_data,
    aes(x = rs1861_C, y = PHENOTYPE, size = count),
    color = "blue"
  ) +

  geom_line(
    data = geno_range_A1,
    aes(x = rs1861_C, y = predicted_prob_A1),
    color = "red",
    size = 1.2
  ) +

  scale_size_continuous(range = c(2, 10)) +
  labs(
    title = "Linear Regression: PHENOTYPE ~ rs1861",
    x = "Genotype (0/1/2)",
    y = "Phenotype ",
    size = "Count"
  ) +

  coord_cartesian(xlim = c(-0.5, 2.5)) +
  theme_minimal()

print(p1)
```

(iv) Convert the genotype coding for rs1861 to recessive coding.

```{r}     
geno_subset_rec <- geno_subset_A1 

geno_subset_rec[, "rs1861_C"] <- as.data.frame(
  lapply(geno_subset_A1[, "rs1861_C"], function(x) ifelse(x == 2, 1, 0))
)
```

(v) Conduct a linear regression to test the association between the recessive-coded rs1861 and the phenotype. What is the p-value?

```{r,message=FALSE, warning=FALSE}
model.rec <- glm(PHENOTYPE ~ rs1861_C, data = geno_subset_rec, family = gaussian)
summary(model.rec)
```

The p-value is <2e-16

(vi) Plot the scatterplot of phenotype versus the recessive-coded genotype of rs1861. Add the regression line to the plot.
       
```{r, message=FALSE, warning=FALSE}         

 geno_range_rec <- data.frame(
   rs1861_C = seq(0, 1, length.out = 100)
 )

 geno_range_rec$predicted_prob_rec <- predict(model.rec, newdata = geno_range_rec, type = "response")

count_data_rec <- geno_subset_rec %>%
  filter(!is.na(rs1861_C), !is.na(PHENOTYPE)) %>%
  group_by(rs1861_C, PHENOTYPE) %>%
  summarise(count = n(), .groups = "drop")

p2=ggplot() +
  geom_point(
    data = count_data_rec,
    aes(x = rs1861_C, y = PHENOTYPE, size = count),
    color = "blue"
  ) +

  geom_line(
    data = geno_range_rec,
    aes(x = rs1861_C, y = predicted_prob_rec),
    color = "red",
    size = 1.2
  ) +

  scale_size_continuous(range = c(2, 10)) +
  labs(
    title = "Linear Regression: PHENOTYPE ~ rs1861 (Recessive)",
    x = "Genotype (0/1/2)",
    y = "Phenotype ",
    size = "Count"
  ) +

  coord_cartesian(xlim = c(-0.5, 2.5)) +
  theme_minimal()

print(p2)
```

(vii) Which model fits better? Justify your answer.

As a lower AIC value means a better fit of model, the additive model with AIC 10988 fits the data better than the recessive model with AIC 11000

### Criteria

| Criteria | Complete | Incomplete |
|------------------------|------------------------|------------------------|
| **Data Inspection** | Correct sample/SNP counts and variable type identified. | Missing or incorrect counts or variable type. |
| **Allele Frequency Estimation** | Correct allele and minor allele frequencies computed. | Frequencies missing or wrong. |
| **Hardy–Weinberg Equilibrium Test** | Correct PLINK command and p-value extraction in R. | PLINK command or extraction incorrect/missing. |
| **Genetic Association Test** | Correct regressions, plots, coding, and interpretation. | Regression, plots, or interpretation missing/incomplete. |

### Submission Information

📌 Please review our [Assignment Submission Guide](https://github.com/UofT-DSI/onboarding/blob/main/onboarding_documents/submissions.md) for detailed instructions on how to format, branch, and submit your work. Following these guidelines is crucial for your submissions to be evaluated correctly.

#### Note:

If you like, you may collaborate with others in the cohort. If you choose to do so, please indicate with whom you have worked with in your pull request by tagging their GitHub username. Separate submissions are required.

------------------------------------------------------------------------

#### Submission Parameters

-   Submission Due Date: `11:59 PM – 16/03/2026`

-   Branch name for your repo should be: `assignment-1`

-   What to submit for this assignment:

    -   Populate this Quarto document (`assignment_1.qmd`).
    -   Render the document with Quarto: `quarto render assignment_1.qmd`.
    -   Submit both `assignment_1.qmd` and the rendered HTML file `assignment_1.html` in your pull request.

-   What the pull request link should look like for this assignment: `https://github.com/<your_github_username>/gen_data/pull/<pr_id>`

    -   Open a private window in your browser. Copy and paste the link to your pull request into the address bar. Make sure you can see your pull request properly. This helps the technical facilitator and learning support team review your submission easily.

------------------------------------------------------------------------

Checklist:

-   Created a branch with the correct naming convention.
-   Ensured that the repository is public.
-   Reviewed the PR description guidelines and adhered to them.
-   Verified that the link is accessible in a private browser window.
-   Confirmed that both `assignment_1.qmd` and `assignment_1.html` are included in the pull request.

If you encounter any difficulties or have questions, please don't hesitate to reach out to our team via our Slack help channel. Our technical facilitators and learning support team are here to help you navigate any challenges.
